Fast response of mesophyll conductance to light availability allows shade leaves to take advantage of sunflecks.

### 2015-08-18 manuscript/manuscript_wtc3.Rmd
started corrections for gm manuscript
did some examples for sachin

---
title: "Response of mesophyll conductance to light availability allows shade leaves to lie in wait"
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: references.bib
---

@@

# Introduction
The need to understand and predict carbon (C) uptake in forest ecosystems is crucially important for assessing the impacts of environmental change. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with increasing leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor], and photosynthesis (A) increases with irradiance from shade to sun until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy C gain it is thus necessary to account for the non-linear response of A to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

\
The distribution of resources required for A, including nitrogen (N) and water, are also partially defined by canopy light gradients. As A has a saturating response with light and a dependence on N, leaf N should be distributed in the upper canopy to maximize canopy C gain [@field1983allocating; @field1986photosynthesis; @peltoniemi2012co]. This can lead to changes in chlorophyll per unit N, chlorophyll a:b ratios, electron transport capacity per unit chlorophyll, and ratios of electron transport capacity to Rubisco activity across leaf types [@evans2001photosynthetic]. Sun leaves also experience greater water limitations in the upper canopy, despite effective vascular systems developed for high radiation loads and transpiration [@sellin2008effects; @niinemets2012optimization]. Higher rates of A and stomatal conductance (g~s~) can only be sustained if the hydraulic conductance (K) is also large enough to avoid low leaf water potentials [@hubbard2001stomatal]. Optimal photosynthetic N investment in the upper canopy will be wasted if photosynthetic capacity exceeds hydraulic supply [@niinemets2012optimization], thus K should also be higher in the upper canopy to supply sunlit leaves with sufficient water [@burgess2006hydraulic;@sellin2007effects ;@sellin2008effects].

\
These theoretical distributions of N and water regulate leaf physiological traits that constrain A differently for sun and shade leaves. The balance between the costs and energetic benefits of these leaf traits affecting photosynthetic capacity create trade-offs impacting canopy C gain [@givnish1988adaptation]. During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using] and trade-offs exists that constrain each of these diffusion pathways. For optimal leaf C gain these pathways should act to minimize the resistance to CO~2~ diffusion while also limiting the intrinsic energy, water and construction costs which diminish the C return from A. As these processes should not be considered independent of each other and must be integrated with hydraulic constraints [@griffiths2013you] predicting their behaviour within canopies remains difficult.

\
First, g~s~ should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. This is because stomata are hypothesized to exhibit an optimal behaviour to maximize C gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Second, variation in photosynthetic capacity between sun and shade leaves has also been hypothesized to be due to differences in g~m~ [@duursma2006vertical]. Mesophyll conductance can impose limitations on A as large as those associated with g~s~ [@warren2008stand; @ubierna2011estimation] and these limitations can reduce the efficiency of N use in A [@niinemets2007photosynthesis]. Interpreting the coupled responses of both g~s~ and g~m~ to A across sun and shade leaves has major implications for predicting canopy C gain, but empirical measurements across tree canopies are still lacking.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by the slow response of stomatal opening, effectively limiting the maximum assimilation rate that can be achieved [@way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both long and rapid time scales, possibly faster than g~s~ [@flexas2008mesophyll]. However, anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are unlikely to adapt during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
1.	If whole tree canopies are optimized for C gain, then leaf N, K and photosynthetic capacity were predicted to be higher in full sun leaves.

\
2.	Under comparable leaf VPD, g~s~ and A should be proportional across both sun and shade leaves, while g~m~ was predicted to scale positively with photosynthetic capacity.

@@

\
4.	The effects of climate warming was predicted to be greater in sun than shade leaves, seen as increases in stomatal closure and reductions in leaf C gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment of +3&deg;C (ET, Figure `r airvarsfig`). Due to the temperature treatment air humidity was not explicitly controlled. The CO~2~ concentration inside the chamber was set to match outside air, tracking diurnal changes.  Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported here.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  The top soils at this site, used in both pots and chambers, are an alluvial formation of low-fertility sandy loam soils with low organic matter and low water holding capacity. After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored. Gas exchange measurements were initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential.
\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held quantum sensor (LI-COR, Lincoln, NE, USA) was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the local light environment of full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

@@

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials (MPa) were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). Predawn leaf water potential was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E, mmol^-2^ s^-1^) from gas exchange were then used to calculate leaf specific hydraulic conductance (K) through the equation:

@@

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous transpiration efficiency (ITE) was calculated as leaf photosynthesis divided by transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear regression, where K=0.5 [see @duursma2013near].

##Data analysis
Differences in experimental parameters to either the warming treatment or leaf type were analysed by mixed-effects models in R with WTC as a random effect. Explained variance (R^2^) of mixed models were computed as in  [@nakagawa2013general]. Confidence intervals (95 %) of mixed effect linear models of A as a function of different physiological parameters were generated using bootstrapping methods with 999 simulations in the 'lme4' package [@bates2013lme4] in R. For non-linear mixed models models confidence intervals were estimated by fitting a generalized additive model to the data with the 'mgcv' package in R.  Results were considered significant at P &leq; 0.05.

@@

## Leaf resource distribution
Across all measurement campaigns 1 m canopy integrated PPFD was significantly different between sun and shade leaves (P<0.001) and PPFD was reduced by 78% in the shade (Figure `r parfig`). The distribution of H~2~O, via K, was not different across sun and shade leaves (Table 1).  This was the result of neither $\Psi_{pd}$, $\Psi_l$ (Table 1) nor E (Table 2) differing between leaf types. Leaf N, on an area basis, was approximately 20% higher in sun leaves than in shade leaves (Table1). Leaf mass per area was not different between leaf types. No effect of the warming treatment was detected with  PPFD, $\Psi_{pd}$, $\Psi_l$, K, E, N~area~ or LMA either within or across leaf types.

## Leaf photosynthesis rates and photosynthetic capacity
Mean leaf photosynthesis rates were significantly higher in sun compared to shade leaves (+23%), under their local light environment (Table 2). Following an increase in available light, A of shade leaves was significantly greater than both shade leaves at low light and sun leaves (P < 0.001). No effect of the warming treatment was detected on rates of A of sun leaves or shade leaves at low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

\
The photosynthetic parameters J~max~ and Vc~max~ were both significantly higher in sun than shade leaves (Table 1), as evident from AC~i~ curves from each chamber (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Within individual chambers Vc~max~ was positively related to leaf N~area~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b).  Additionally, leaf N~area~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c).  As Vc~max~ is considered implicit with g~m~ (results below) with conventional AC~i~ curves, a AC~c~ curves were also simulated to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~.  When accounting for the effect of g~m~, photosynthetic capacity was still greater in sun leaves than shade leaves (Figure `r accfig`)

## Leaf water use efficiency
Leaf ITE was significantly greater in sun leaves than in shade leaves at low light (P = 0.001). Following an increase in available light, ITE of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001).  ITE in sun leaves was reduced in the warming treatment but no effect was noted in shade leaves with low or high light (Table 2). For all leaf types and light treatments there was a strong response of ITE to VPD, and individual data points broadly corresponded to response curves from the optimal ITE model (Figure `r waterusefig`a).  The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04.  Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a)

\
Leaf $\delta$^13^C, as an index of water use efficiency [@marshall2007sources], significantly decreased from sun leaves to shade leaves by ca. 2&permil; (Table 1). No effects of the warming treatment on leaf $\delta$^13^C were detected. Leaf $\delta$^13^C and N~area~ were positively correlated for all leaves (P<0.001, Figure `r waterusefig`b), with less negative $\delta$^13^C and higher N investment in sun leaves.

## Leaf physiology
Mean g~s~ was significantly higher in shade compared to sun leaves (+18%) under their local light environment (P = 0.004). Increases in g~s~ led to increases in A in sun leaves, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves ( Figure `r gebarfig`a). With increased light availability, both increases in A and g~s~ in shade leaves exhibited similar behavior as observed in sun leaves.

@@

\
Mean g~m~ was higher in sun compared to shade leaves (+27%) under their local light environment (P < 0.001).  Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves (Table 2). Relationships between g~m~ and A were nearly proportional within leaf types and light treatments (Figure `r gebarfig`b,c). Photosynthesis scaled positively with increases in g~m~ for all leaves and the large increases in g~m~ in shade leaves under high light resulted in the highest rates of A (Figure `r agmgsfig`b). No differences in g~s~ nor g~m~ were detected with the warming treatment within leaf types. Additionally, neither g~s~ nor g~m~ varied significantly across measurements campaigns (Figures `r get1fig`b and `r get2fig`b, respectively) and a slightly negative relationship with g~m~ and increasing leaf temperature was detected.

\
The combination of a non-limiting supply of leaf K and more open stomata lead to a significant increase in C~i~ in shade compared to sun leaves (Figure `r ciccfig`a). As increases in A in shade leaves were associated with higher rates of g~s~ the C~i~ was similar than under low light but still greater than sun leaves. As the drop in CO~2~ for intecellular spaces to the cholorplast, C~i~-C~c~, measures the coordination between g~m~ and A [@caemmerer2014temperature] an increase in this drawdown was detected in shade leaves at high light (Figure `r ciccfig`b,c).  This was a direct result of the increase in g~m~ and its subsequent direct effect on A, when compared to both sun and shade leaves that are acclimated to their local light environment. CO~2~ drawdown from C~a~ to C~i~ and C~i~ to C~c~ were both relatively stable across the range of temperatures measured and gas exchange campaigns (Figure `r get2fig`c and `r get1fig`c, respectively).

@@

## Resource distribution and photosynthetic capacity
The allocation of N constrains photosynthesis rates and is thus a key trait in determining the relative contribution of individual leaves to canopy C gain.  Decreasing light availability should decrease the investment into photosynthetic enzyme within a canopy (Mooney and Gulmon 1979).  As a result, acclimation of photosynthetic capacity to irradiance is typically reflected in the key photosynthetic biochemical parameters Vc~max~ and J~max~ [@farquhar1980biochemical]. Our data agree with these conventional conclusions as the distribution of N, both biochemical parameters and A were all reduced in shade leaves.

@@

\
This behaviour may be explained by the response of shade leaves to increases in light availability. Contrary to hypothesis, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite canopy light extinction [@pearcy1990sunflecks]. Unfortunately, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy.

>>>>

---
title: "Fast response of mesophyll conductance to light availability allows shade leaves to take advantage of sunflecks."
author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
  pdf_document: default
  html_document:
    number_sections: yes
csl: new-phytologist.csl
bibliography: references.bib
---

@@

# Introduction
Understanding and predict carbon (C) uptake in forest ecosystems is crucially important for assessing the impacts of environmental change. Specifically, this involves accurate upscaling of leaf level measurements to predict whole canopy function. Due to the costs and limitations of efficient light harvesting, plants cannot expose all leaves to full sun [@niinemets2010review], making simple upscaling problematic. Incident PPFD declines exponentially with cumulative leaf area index, creating a steep light gradient from the canopy top to bottom [@monsi2005factor] and photosynthesis (A) responds to this changing irradiance until until biochemical limitation occurs [@evans1995carbon]. Consequently, leaves modify structural and functional properties to efficiently intercept variable light and enhance photosynthetic capacity [@vogelman1996leaves; @valladares2004photosynthetic]. To estimate whole canopy C gain it is thus necessary to account for the non-linear response of A to light by distinguishing between shaded and sunlit leaves [@de1997simple; @linderson2012up].

\
The distribution of resources required for A, including nitrogen (N) and water, are also partially defined by canopy light gradients. As A has a saturating response with light and a dependence on N, it has been argued that leaf N should be proportional to PPFD along the canopy gradient to maximize canopy C gain at a given total canopy N [@field1983allocating; @field1986photosynthesis; @peltoniemi2012co]. This can lead to changes in chlorophyll per unit N, chlorophyll a:b ratios, electron transport capacity per unit chlorophyll, and ratios of electron transport capacity to Rubisco activity across leaf types [@evans2001photosynthetic]. Sun leaves frequently experience greater water limitations in the upper canopy, despite effective vascular systems developed for high radiation loads and transpiration [@sellin2008effects; @niinemets2012optimization]. Higher rates of A and stomatal conductance (g~s~) can only be sustained if the hydraulic conductance (K) is also large enough to avoid low leaf water potentials [@hubbard2001stomatal]. Optimal photosynthetic N investment in the upper canopy will be wasted if photosynthetic capacity exceeds hydraulic supply [@niinemets2012optimization; @peltoniemi2012co], thus K should also be higher in the upper canopy to supply sunlit leaves with sufficient water [@burgess2006hydraulic;@sellin2007effects ;@sellin2008effects].

\
These theoretical distributions of N and water regulate leaf physiological traits that constrain A differently for sun and shade leaves. The balance between the costs and benefits of these leaf traits affecting photosynthetic capacity create trade-offs impacting canopy C gain [@givnish1988adaptation]. During A, CO~2~ diffuses from the atmosphere through stomata (g~s~), intercellular air space and into the chloroplast for fixation by Rubisco (mesophyll conductance, g~m~) [@tazoe2011using]. Trade-offs exists that constrain each of these diffusion pathways, although these are yet to be explictily quantified for g~m~. For optimal leaf C gain these pathways should act to minimize the resistance to CO~2~ diffusion while also limiting the intrinsic energy, water and construction costs which diminish the C return from A. As these processes should not be considered independent of each other and must be integrated with hydraulic constraints [@griffiths2013you] predicting their behaviour within canopies remains difficult.

\
First, g~s~ should be efficiently distributed within a canopy to utilize supplies of light, N and water to maximize A [@peltoniemi2012co]. This is because stomata are hypothesized to exhibit an optimal behaviour to maximize C gain while simultaneously minimizing water loss through transpiration [@cowan1977stomatal]. Second, variation in photosynthetic capacity between sun and shade leaves has also been hypothesized to be due to differences in g~m~ [@piel2002effect; @duursma2006vertical; @warren2007internal]. Mesophyll conductance can impose limitations on A as large as those associated with g~s~ [@warren2008stand; @ubierna2011estimation] and these limitations can reduce the efficiency of N use in A [@niinemets2007photosynthesis]. Interpreting the coupled responses of both g~s~ and g~m~ to A across sun and shade leaves has major implications for predicting canopy C gain, but empirical measurements across tree canopies are still lacking.

\
Additionally, assessing shade leaf behaviour is made difficult with accounting of short term light fluctuations within a canopy, via sunflecks. How shade leaves utilize sunflecks for short term C gain depends on the combined response time of g~s~ and g~m~ and the underlying photosynthetic biochemistry acclimated to a low light environment [@pearcy1990sunflecks]. For example, at short timescales the utilization of sunflecks is limited by delayed responses of stomata, effectively limiting the maximum assimilation rate that can be achieved [@vico2011effects; @way2012sunflecks]. Mesophyll conductance has been shown to respond to environmental factors across both short and long time scales, possibly faster than g~s~ [@flexas2008mesophyll]. However, anatomical parameters which regulate g~m~ with changing irradiance such as chloroplast surface area [@terashima2006irradiance] and mesophyll thickness [@boardman1977comparative] are unlikely to adapt during short light fluctuations. The physiological behaviour of shade leaves to maximize C gain must be assessed as both a degree of acclimation to local irradiance and as a potential response to transitory light availability.

@@

\
1.	If whole tree canopies are optimized for C gain, then leaf N, K and photosynthetic capacity were predicted to be higher in sun leaves compared to shade leaves.

\
2.	Under comparable leaf VPD, g~s~ should be proportional to A across both sun and shade leaves and g~m~ should scale positively with photosynthetic capacity.

@@

\
4.	The effects of climate warming were predicted to be greater in sun than shade leaves, seen as a decrease in g~s~ and leaf C gain during summer months.

@@

## Whole tree chamber experimental design
Twelve *Eucalyptus tereticornis* seedlings, chosen from a single local Cumberland plain cohort, were planted in March 2013 into 12 whole-tree chambers (WTC) at the Hawkesbury Forest Experiment site near Richmond, NSW, Australia. Each chamber has a maximum height of 10 m and seedlings were grown for 15 months.  A detailed description of the WTC operation and design is available in [@barton2010whole]. Six chambers were set to match outside ambient air temperatures (AT) while the remaining 6 experienced a constant elevated air temperature treatment of +3&deg;C (ET, Figure `r airvarsfig`). Trees were watered weekly with 70 L from March 2013 to November 2013.  From December 2013 to final harvest trees were watered fortnightly with the mean monthly (100 yr) rainfall amount.  In February 2013 half of the chambers (3 each of AT and ET) were subjected to a drought treatment by withholding watering. Due to a limited range of data for the drought treatment only well-watered trees are reported, which reduces the sample size of WTC (n=6) after the initiation of this treatment.

\
Before seedlings were planted into each chamber they were maintained under well watered conditions in 35 L pots and kept inside each chamber. This allowed for seedlings to gain sufficient biomass before planting while also allowing them to acclimate to chamber temperature treatments. Seedlings were planted into each chamber after mean seedling height reached 100 cm.  The top soils at this site, used in both pots and chambers, are an alluvial formation of low-fertility sandy loam soils with low organic matter and low water holding capacity. A root exclusion barrier extended from chamber walls to the hard layer (ca. 1 m) and roots were allowed to grow freely below the barrier. After 2 months, floors were installed 45 cm from the soil surface.  This enabled chamber fluxes of CO~2~ and H~2~O from the whole tree canopy to be monitored (for methods see Barton et al. [-@barton2010whole]). Leaf gas exchange measurements were then initiated in October 2013 when trees had both ample height growth and canopy development for realistic canopy light gradients to be measured.  At this point, trees under AT treatment had a mean diameter of 28.2&pm;1.1 mm, height of 348&pm;15.1 cm and an estimated leaf area of 3.9&pm;0.1 m^2^. For ET treatments trees had a mean diameter of 34.1&pm;2.1 mm, height of 418.3&pm;23.1 cm and an estimated leaf area of 6.2&pm;0.2 m^2^.

##Leaf gas exchange, online carbon isotope discrimination and mesophyll conductance
Leaf gas exchange measurements were performed monthly through to the final harvest.  Measurements were taken on a representative sun and shade leaf for each tree during each measurement campaign.  The newest fully expanded leaf from the stem apex was chosen for gas exchange measurements and sun leaves were measured in the upper third of the canopy. In order to minimize any confounding effects of leaf age, shade leaves were measured first in the lower canopy then gradually higher during each campaign.  As shade leaves most likely developed slower this assured that older leaves in the lower canopy were avoided. The nearest leaf on each branch was sampled for measurement of predawn leaf water potential.

\
Prior to gas exchange measurements photosynthetic photon flux density (PPFD) was recorded at the individual leaf level and at the canopy position for each selected leaf.  A hand-held quantum sensor (LI-COR, Lincoln, NE, USA) was used to record leaf level PPFD to ensure that chosen leaves were positioned in the desired light environment, either full sun or full shade. A ceptometer (AccuPAR LP-80, Decagon Devices, Pullman, USA) was then used to measure 1 m integrated PPFD at the canopy height of each chosen leaf type. Five ceptometer readings were recorded within the canopy at random locations of the height of each selected leaf.  The mean of these readings was assumed to represent the local light environment of full sun and shade leaves for each tree. All measurements of PPFD and gas exchange were performed on full sun days between 10:00-14:30 h.

@@

\
Predawn ($\Psi_{pd}$) and midday ($\Psi_l$) leaf water potentials (MPa) were measured for sun and shade leaves during each gas exchange campaign using a PMS 1505D pressure chamber (PMS Instruments, Albany, OR, USA). Predawn leaf water potential was measured before sunrise on the same day as gas exchange measurements. Leaves used for gas exchange were immediately sampled for $\Psi_l$ once measurements were completed. All leaves were detached and immediately stored inside foil covered bags before water potential measurements were performed. Leaf water potential and transpiration (E, mmol^-2^ s^-1^) from gas exchange were then used to calculate leaf-specific hydraulic conductance (K) through the equation:

@@

where the term Hp~w~g represents the gravitational potential assumed to be minimal in this experiment due to small tree size [@whitehead1998regulation]. Leaf level instantaneous water use efficiency (WUE~i~) was calculated as leaf photosynthesis divided by transpiration. The g~1~ parameter was estimated from ITE to VPD response curves by fitting a rearranged optimal stomatal conductance model for ITE [@medlyn2012reconciling] using non-linear regression, where K=0.5 [see @duursma2013near].

##Data analysis
Differences in experimental parameters to either the warming treatment or leaf type were analysed by mixed-effects models in R with WTC as a random effect. Explained variance (R^2^) of mixed models were computed as in  [@nakagawa2013general]. Confidence intervals (95 %) of mixed effect linear models of A as a function of different physiological parameters were generated using bootstrapping methods with 999 simulations in the 'lme4' package [@bates2013lme4] in R. For non-linear relationships, confidence intervals were estimated by fitting a generalized additive model to the data with the 'mgcv' package in R, using WTC as a random effect.  Results were considered significant at P &leq; 0.05.

@@

## Leaf resource distribution
Across all measurement campaigns 1 m canopy integrated PPFD was significantly different between sun and shade leaves (P<0.001) and PPFD was reduced by 78% in the shade (Figure `r parfig`). Leaf-specific hydraulic conductance was not different across sun and shade leaves (Table 1).  This was the result of neither $\Psi_{pd}$, $\Psi_l$ (Table 1) nor E (Table 2) differing between leaf types. Leaf N, on an area basis, was approximately 20% higher in sun leaves than in shade leaves (Table1). Leaf mass per area was not different between leaf types. No effect of the warming treatment was detected with  PPFD, $\Psi_{pd}$, $\Psi_l$, K, E, N~area~ or LMA either within or across leaf types.

## Photosynthetic capacity and leaf photosynthesis rates 
The photosynthetic parameters J~max~ and Vc~max~ were both significantly higher in sun than shade leaves (Table 1), as evident from AC~i~ curves from each chamber (Figure `r acifig`a). Within leaf types, no effect of the warming treatment was detected on either parameter. Within individual chambers Vc~max~ was positively related to leaf N~area~ across leaf types and temperature treatments (P = 0.01, Figure `r acifig`b). Additionally, leaf N~area~ was positively related to A across gas exchange campaigns and leaf types (P < 0.001, Figure `r acifig`c). As g~m~ (results below) is included in the estimates of J~max~ and Vc~max~ with conventional AC~i~, AC~c~ curves were also simulated to determine if treatment differences in AC~i~ parameters where instead the result of differences in g~m~. When accounting for the effect of g~m~, chloroplastic Vc~max~ and J~max~ were still higher in sun leaves (Figure `r accfig`).

\
Mean leaf photosynthesis rates were significantly higher in sun compared to shade leaves (+23%), when measured at their local light environment (Table 2). Following an increase in available light, A of shade leaves was significantly greater than both shade leaves at low light and sun leaves (P < 0.001). No effect of the warming treatment was detected on rates of A of sun leaves or shade leaves at low or high light. Photosynthesis within leaf types and treatments was similar through time and across the range of leaf temperatures measured (Figure `r get1fig`a).

## Leaf water use efficiency
Leaf WUE~i~ was significantly greater in sun leaves than in shade leaves at low light (P = 0.001). Following an increase in available light, WUE~i~ of shade leaves did not differ from shade leaves at low light and was still significantly lower than sun leaves (P < 0.001). WUE~i~ in sun leaves was reduced in the warming treatment but no effect was noted in shade leaves with low or high light (Table 2). For all leaf types and light treatments there was a strong response of WUE~i~ to VPD, and individual data points broadly corresponded to response curves from the optimal WUE~i~ model (Figure `r waterusefig`a). The mean estimated g~1~ for sun leaves was 1.51&pm;0.11 and for shade leaves with low and high light was 2.59&pm;0.12 and 2.74&pm;0.04. Within leaf types and light treatments the response of VPD to leaf temperature was similar across all measurement campaigns (Figure `r get2fig`a)

\
Leaf $\delta$^13^C, as an index of integrated water use efficiency [@marshall2007sources], significantly decreased from sun leaves to shade leaves by ca. 2&permil; (Table 1). No effects of the warming treatment on leaf $\delta$^13^C were detected. Leaf $\delta$^13^C and N~area~ were positively correlated for all leaves (P<0.001, Figure `r waterusefig`b), with less negative $\delta$^13^C and higher N investment in sun leaves.

## Leaf physiology
Mean g~s~ was significantly higher in shade compared to sun leaves (+18%) under their local light environment (P = 0.004). Increases in g~s~ were postively correlated to increases in A in sun leaves, however, g~s~ and A were not correlated in shade leaves under low light (Figure `r agmgsfig`a). Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves ( Figure `r gebarfig`a). With increased light availability, both increases in A and g~s~ in shade leaves exhibited similar behavior as observed in sun leaves.

@@

\
Mean g~m~ was higher in sun compared to shade leaves (+27%) under their local light environment (P < 0.001).  Following an increase in available light, g~s~ of shade leaves was significantly greater than both shade leaves at low light and sun leaves (Table 2). Mesophyll conductance was nearly proportional to A within leaf types and light treatments (Figure `r gebarfig`b,c). Photosynthesis scaled positively with increases in g~m~ for all leaves and the large increases in g~m~ in shade leaves under high light resulted in the highest rates of A (Figure `r agmgsfig`b). No differences in g~s~ nor g~m~ were detected with the warming treatment within leaf types. Additionally, neither g~s~ nor g~m~ varied significantly across measurements campaigns (Figure `r get1fig`b and `r get2fig`b, respectively) and only a weak negative relationship with g~m~ and increasing leaf temperature was detected (Figure `r get1fig`b).

\
The combination of a non-limiting supply of leaf K and more open stomata lead to a significant increase in C~i~ in shade compared to sun leaves (Figure `r ciccfig`a). As increases in A in shade leaves were associated with higher rates of g~s~ the C~i~ was similar than under low light but still greater than sun leaves. As the drop in CO~2~ for intercellular spaces to the cholorplast, C~i~-C~c~, measures the coordination between g~m~ and A [@caemmerer2014temperature] an increase in this drawdown was detected in shade leaves at high light (Figure `r ciccfig`b,c).  This was a direct result of the increase in g~m~ and its subsequent direct effect on A, when compared to both sun and shade leaves that are acclimated to their local light environment. CO~2~ drawdown from C~a~ to C~i~ and C~i~ to C~c~ were both relatively stable across the range of temperatures measured and gas exchange campaigns (Figure `r get2fig`c and `r get1fig`c, respectively).

@@

## Resource distribution and photosynthetic capacity
The allocation of N~a~ constrains photosynthesis rates and is thus a key trait in determining the relative contribution of individual leaves to canopy C gain.  Decreasing light availability should decrease the investment into photosynthetic enzyme within a canopy (Mooney and Gulmon 1979).  As a result, acclimation of photosynthetic capacity to irradiance is typically reflected in the key photosynthetic biochemical parameters Vc~max~ and J~max~ [@farquhar1980biochemical]. Our data agree with these conventional conclusions as the distribution of N~a~ , both biochemical parameters and A were all reduced in shade leaves.

@@

\
This behaviour may be explained by the response of shade leaves to increases in light availability. Contrary to hypothesis, shade leaves were able to increase A to rates equivalent or even outperforming sun leaves under identical light intensity. It is possible that sustaining higher g~s~, at the cost of larger transpiration losses, is a strategy to increase the utilization efficiency of sunflecks. These results should not be entirely unexpected as shade leaves experience interchanging periods of sun and shade, despite lower mean PPFD [@pearcy1990sunflecks]. Unfortunately, empirical studies that assess photosynthetic responses to sunflecks generally focus on seedlings [@kuppers1993leaf;@leakey2002relative] and understory plants [@chazdon1991importance; @brantley2009contribution]. Our findings highlight a critical need for empirical measurements of shade leaves under dynamic light environments in order to accurately scale C gain from leaf to canopy.
